# STAT 165 Final Project

## *Southwest Airlines Baggage Fees Effects on Passenger Volume*

## *Forecast for June 2025*

### Group Members: Ari Hadjiyianni, Assmaa Aziz, Evan Vlahos

```{r}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(boot)

options(warn=0)
```

## Fermi Estimation
By Evan Vlahos

### Load all data
```{r}
# Vector that contains the major airlines that we have data for
airlines = c("Alaska", "Frontier", "JetBlue", "Southwest", "Spirit")
months = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
# Now setup the translation vector - essentially a lookup table
names(months) <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12)
seasons = c("Winter", "Winter", "Spring", "Spring", "Spring", "Summer", "Summer", "Summer", "Fall", "Fall", "Fall", "Winter")
names(seasons) <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12)
quarter_str = c("Q1", "Q2", "Q3", "Q4")
quarters = c(1,2,3,4)
```

```{r}
suppressWarnings({
  
# Reading Flight Data
flights_data = data.frame()
# Looping through our airlines vector to extract the flight data
for(i in 1:length(airlines))
{
  # Making the path to the file for each of the airlines
  airline_file = paste("US-Airline-Flights/",airlines[i],"Flights.csv", sep="")
  # Read the file
  df = read.csv(airline_file)
  # Remove the incomplete years from the dataframe
  df <- df %>%
    filter((Year > 2002 & Year < 2020) | (Year > 2022))
  # Make a column for the Airline string
  df["Airline"] = airlines[i]
  df$MonthName <- months[as.character(df$Month)]
  df$MonthName <- factor(df$MonthName, levels = months)
  df$Season <- seasons[as.character(df$Month)]
  df$Year <- as.integer(gsub(",", "", df$Year))
  df$TOTAL <- as.numeric(gsub(",", "", df$TOTAL))
  df$DOMESTIC <- as.numeric(gsub(",", "", df$DOMESTIC))
  df$INTERNATIONAL <- as.numeric(gsub(",", "", df$INTERNATIONAL))
  # Append the df dataframe to the flights_data dataframe
  flights_data = rbind(flights_data, df)
}

})
```

```{r}
# Reading Passenger Data
passengers_data = data.frame()
# Looping through our airlines vector to extract the passenger data
for(i in 1:length(airlines))
{
  # Making the path to the file for each of the airlines
  airline_file = paste("US-Airline-Passengers/",airlines[i],"Passengers.csv", sep="")
  # Read the file
  df = read.csv(airline_file)
  # Remove the incomplete years from the dataframe
  df <- df %>%
    filter((Year > 2002 & Year < 2020) | (Year > 2022))
  # Make a column for the Airline string
  df["Airline"] = airlines[i]
  df$MonthName <- months[as.character(df$Month)]
  df$MonthName <- factor(df$MonthName, levels = months)
  df$Season <- seasons[as.character(df$Month)]
  df$TOTAL <- as.numeric(gsub(",", "", df$TOTAL))
  df$DOMESTIC <- as.numeric(gsub(",", "", df$DOMESTIC))
  df$INTERNATIONAL <- as.numeric(gsub(",", "", df$INTERNATIONAL))
  # Append the df dataframe to the passengers_data dataframe
  passengers_data = rbind(passengers_data, df)
}

```

```{r}
# Reading Available Seat Miles Data
available_seat_miles_data = data.frame()
# Looping through our airlines vector to extract the available seat mile data
for(i in 1:length(airlines))
{
  # Making the path to the file for each of the airlines
  airline_file = paste("US-Airline-Available-Seat-Miles/",airlines[i],"_ASM.csv", sep="")
  # Read the file
  df = read.csv(airline_file)
  # Remove the incomplete years from the dataframe
  df <- df %>%
    filter((Year > 2002 & Year < 2020) | (Year > 2022))
  # Make a column for the Airline string
  df["Airline"] = airlines[i]
  df$MonthName <- months[as.character(df$Month)]
  df$MonthName <- factor(df$MonthName, levels = months)
  df$Season <- seasons[as.character(df$Month)]
  df$Year <- as.integer(gsub(",", "", df$Year))
  df$TOTAL <- as.numeric(gsub(",", "", df$TOTAL))
  df$DOMESTIC <- as.numeric(gsub(",", "", df$DOMESTIC))
  df$INTERNATIONAL <- as.numeric(gsub(",", "", df$INTERNATIONAL))
  # Append the df dataframe to the flights_data dataframe
  available_seat_miles_data = rbind(available_seat_miles_data, df)
}
 
```

```{r}
# Reading Net Income Data
net_income_data = data.frame()
# Looping through our airlines vector to extract the net income data
for(i in 1:length(airlines))
{
  # Making the path to the file for each of the airlines
  airline_file = paste("US-Airline-Net-Income/",airlines[i],"_NI.csv", sep="")
  # Read the file
  df = read.csv(airline_file)
  # Remove the incomplete years from the dataframe
  df <- df %>%
    filter((Year > 2002 & Year < 2020) | (Year > 2022))
 # Make a column for the Airline string
  df["Airline"] = airlines[i]
  df$Quarter <- factor(df$Quarter, levels = quarters)
  df$Year <- as.integer(gsub(",", "", df$Year))
  df$TOTAL <- as.numeric(gsub(",", "", df$TOTAL))
  df$DOMESTIC <- as.numeric(gsub(",", "", df$DOMESTIC))
  df$LATIN.AMERICA <- as.numeric(gsub(",", "", df$LATIN.AMERICA))
  df$ATLANTIC <- as.numeric(gsub(",", "", df$ATLANTIC))
  df$PACIFIC <- as.numeric(gsub(",", "", df$PACIFIC))
  # Append the df dataframe to the flights_data dataframe
  net_income_data = rbind(net_income_data, df)
}

```

```{r}
# Reading Operating Revenue Data
operating_revenue_data = data.frame()
# Looping through our airlines vector to extract the net income data
for(i in 1:length(airlines))
{
  # Making the path to the file for each of the airlines
  airline_file = paste("US-Airline-Operating-Revenue/",airlines[i],"_OR.csv", sep="")
  # Read the file
  df = read.csv(airline_file)
  # Remove the incomplete years from the dataframe
  df <- df %>%
    filter((Year > 2002 & Year < 2020) | (Year > 2022))
  # Make a column for the Airline string
  df["Airline"] = airlines[i]
  df$Quarter <- factor(df$Quarter, levels = quarters)
  df$Year <- as.integer(gsub(",", "", df$Year))
  df$TOTAL <- as.numeric(gsub(",", "", df$TOTAL))
  df$DOMESTIC <- as.numeric(gsub(",", "", df$DOMESTIC))
  df$LATIN.AMERICA <- as.numeric(gsub(",", "", df$LATIN.AMERICA))
  df$ATLANTIC <- as.numeric(gsub(",", "", df$ATLANTIC))
  df$PACIFIC <- as.numeric(gsub(",", "", df$PACIFIC))
  # Append the df dataframe to the flights_data dataframe
  operating_revenue_data = rbind(operating_revenue_data, df)
}
 
```

```{r}
# Reading Revenue Passenger Miles Data
revenue_passenger_miles_data = data.frame()
# Looping through our airlines vector to extract the load factor data
for(i in 1:length(airlines))
{
  # Making the path to the file for each of the airlines
  airline_file = paste("US-Airline-Revenue-Passenger-Miles/",airlines[i],"_RPM.csv", sep="")
  # Read the file
  df = read.csv(airline_file)
  # Remove the incomplete years from the dataframe
  df <- df %>%
    filter((Year > 2002 & Year < 2020) | (Year > 2022))
  # Make a column for the Airline string
  df["Airline"] = airlines[i]
  df$MonthName <- months[as.character(df$Month)]
  df$MonthName <- factor(df$MonthName, levels = months)
  df$Season <- seasons[as.character(df$Month)]
  df$TOTAL <- as.numeric(gsub(",", "", df$TOTAL))
  df$DOMESTIC <- as.numeric(gsub(",", "", df$DOMESTIC))
  df$INTERNATIONAL <- as.numeric(gsub(",", "", df$INTERNATIONAL))
  # Append the df dataframe to the passengers_data dataframe
  revenue_passenger_miles_data = rbind(revenue_passenger_miles_data, df)
}

```

### Create plots to visualize data
```{r}
# Number of passengers per airline per year grouped by the airlines
ggplot(data = passengers_data, aes(x = MonthName, y = TOTAL/1000000, group = Year)) + 
    geom_line(aes(color=as.character(Year))) +
    facet_wrap(~Airline) +
    labs(title="Number of Passengers (ex-COVID)",
        x ="Year", y = "Total (in millions)") +
    theme(axis.text.x = element_text(angle=90))

```

```{r}
# Annual passengers
annual_pass_data <- passengers_data %>% group_by(Year, Airline) %>% summarize(AnnualPassengers = sum(TOTAL))
```

```{r}
grph <- annual_pass_data %>%  
  filter(Year >= 2006 & Year <= 2019) %>%
  ggplot(aes(x = Year, y = AnnualPassengers / 1000000, group = Airline)) + 
    geom_line(aes(color=Airline)) +
    facet_wrap(~Airline) +
    labs(title="2006-2019 Annual Passenger Count",
        x ="Year", y = "Total (in millions)") + 
    theme(axis.text.x = element_text(angle=90))

print(grph)
```

### June data for all airlines

```{r}
# data for June
month_value = 6 

passengers_data %>% filter(Month == month_value ) %>% ggplot(aes(x = Year, y = TOTAL/1000000, group = Airline)) + 
    geom_line(aes(color=Airline)) +
    facet_wrap(~Airline) +
    labs(title=paste(months[month_value], " Passengers per year",sep=""),
        x ="Year", y = "Total (in millions)") + 
    theme(axis.text.x = element_text(angle=90))

```

```{r}
# Calculating passengers per flight
passenger_dataset <- passengers_data[c("Year", "Month", "Airline", "TOTAL")]
flight_dataset <- flights_data[c("Year", "Month", "Airline", "TOTAL")]
pass_flight <- inner_join(passenger_dataset, flight_dataset, by = c("Year", "Month", "Airline"), suffix = c(".passengers", ".flights"))
pass_flight$pass_flight_ratio = pass_flight$TOTAL.passengers / pass_flight$TOTAL.flights
```

```{r}
# data for June
month_value = 6

grph <- pass_flight %>% filter(Month == month_value ) %>% ggplot(aes(x = Year, y = pass_flight_ratio, group = Airline)) + 
    geom_line(aes(color=Airline)) +
    facet_wrap(~Airline) +
    labs(title=paste("Passengers per Flight for ", months[month_value], sep=""),
        x ="Year", y = "Total") + 
    theme(axis.text.x = element_text(angle=90))

print(grph)

```

```{r}
# available seat miles
month_value = 6

grph <- available_seat_miles_data %>% filter(Month == month_value ) %>% ggplot(aes(x = Year, y = TOTAL/1000000, group = Airline)) + 
    geom_line(aes(color=Airline)) +
    facet_wrap(~Airline) +
    labs(title=paste("Available Seat Miles for ", months[month_value], sep=""),
        x ="Year", y = "Total (in millions)") + 
    theme(axis.text.x = element_text(angle=90))

print(grph)

```

### Annualized Data
```{r}
# Net Income
annual_net_data <- net_income_data %>% group_by(Year, Airline) %>% summarize(AnnualNet = sum(TOTAL))
```

```{r}
# Plotting the Net Income of each airline
grph <- annual_net_data %>%  
  filter(Year >= 2015) %>%
  ggplot(aes(x = Year, y = AnnualNet / 1000, group = Airline)) + 
    geom_line(aes(color=Airline)) +
    facet_wrap(~Airline) +
    labs(title="Net income since 2015 (ex-COVID)",
        x ="Year", y = "Total (in millions)") + 
    theme(axis.text.x = element_text(angle=90))

print(grph)
```

```{r}
annual_revenue_data <- operating_revenue_data %>% filter(Year <= 2024) %>% group_by(Year, Airline) %>% summarize(AnnualTotal = sum(TOTAL)*1000)
```

```{r}
# Operating revenue 

grph <- annual_revenue_data %>%  ggplot(aes(x = Year, y = AnnualTotal/1000000, group = Airline)) + 
    geom_line(aes(color=Airline)) +
    facet_wrap(~Airline) +
    labs(title="Operating revenue",
        x ="Year", y = "Total (in millions)") + 
    theme(axis.text.x = element_text(angle=90))

print(grph)
```

```{r}
annual_passenger_data <- passengers_data %>% filter(Year <= 2024) %>% group_by(Year, Airline) %>% summarize(AnnualTotal = sum(TOTAL))
```

```{r}
grph <- annual_passenger_data %>%  ggplot(aes(x = Year, y = AnnualTotal / 1000000, group = Airline)) + 
    geom_line(aes(color=Airline)) +
    facet_wrap(~Airline) +
    labs(title="Annual Number of Passengers",
        x ="Year", y = "Number of Passengers (in millions)") + 
    theme(axis.text.x = element_text(angle=90))
print(grph)
```

```{r}
annual_pass_flight_ratio <- pass_flight %>% filter(Year <= 2024) %>% group_by(Year, Airline) %>% summarize(AnnualTotal = mean(pass_flight_ratio))
```

```{r}
grph <- annual_pass_flight_ratio %>%  ggplot(aes(x = Year, y = AnnualTotal, group = Airline)) + 
    geom_line(aes(color=Airline)) +
    facet_wrap(~Airline) +
    labs(title="Annualized Passengers per Flight",
        x ="Year", y = "Passengers per Flight") + 
    theme(axis.text.x = element_text(angle=90))
print(grph)
```

```{r}
annual_revenue_passenger_data <- annual_revenue_data
annual_revenue_passenger_data$Revenue_Per_Passenger <- (annual_revenue_data$AnnualTotal) / (annual_passenger_data$AnnualTotal)
```

```{r}
grph <- annual_revenue_passenger_data %>%  ggplot(aes(x = Year, y = Revenue_Per_Passenger, group = Airline)) + 
    geom_line(aes(color=Airline)) +
    facet_wrap(~Airline) +
    labs(title="Revenue per Passenger",
        x ="Year", y = "Revenue per Passenger (in $)") + 
    theme(axis.text.x = element_text(angle=90))
print(grph)
```

```{r}
annual_asm <- available_seat_miles_data %>% filter(Year <= 2024) %>% group_by(Year, Airline) %>% summarize(AnnualTotal = mean(TOTAL))
```

```{r}
# Plot Annual Available Seat Miles
grph <- annual_asm %>%  ggplot(aes(x = Year, y = AnnualTotal/1000000, group = Airline)) + 
    geom_line(aes(color=Airline)) +
    facet_wrap(~Airline) +
    labs(title="Annualized Available Seat Miles ",
        x ="Year", y = "Available Seat Miles (in millions)") + 
    theme(axis.text.x = element_text(angle=90))
print(grph)
```

```{r}
# Year over Year passengers in June for Southwest - 2023 to 2024
jun_southwest_passengers2024 <- passengers_data %>% filter(Airline == "Southwest" & Year %in% c(2023, 2024) & Month == 6)
jun_southwest_passengers2024
```

```{r}
# Year over Year passengers in June for Southwest - 2014 to 2015
jun_southwest_passengers2015 <- passengers_data %>% filter(Airline == "Southwest" & Year %in% c(2014, 2015) & Month == 6)
jun_southwest_passengers2015
```

```{r}
# Year over Year passengers in June for JetBlue
jun_jetblue_passengers2024 <- passengers_data %>% filter(Airline == "JetBlue" & Year %in% c(2023, 2024) & Month == 6)
jun_jetblue_passengers2024
```

```{r}
# Year over Year passengers in June for JetBlue
jun_jetblue_passengers2015 <- passengers_data %>% filter(Airline == "JetBlue" & Year %in% c(2014, 2015) & Month == 6)
jun_jetblue_passengers2015
```

```{r}
# Calculating JetBlue's Growth Factor from 2014 to 2015
jetblue_growth = ((jun_jetblue_passengers2015$TOTAL[2] / jun_jetblue_passengers2015$TOTAL[1]) - 1) * 100
print(paste0("JetBlue's Growth Rate (2014-2015): ", round(jetblue_growth, digits = 2), "%", sep=""))
```

```{r}
# Calculating JetBlue's Growth Factor from 2023 to 2024
jetblue_growth2 = ((jun_jetblue_passengers2024$TOTAL[2] / jun_jetblue_passengers2024$TOTAL[1]) - 1) * 100
print(paste0("JetBlue's Growth Rate (2023-2024): ", round(jetblue_growth2, digits = 2), "%", sep=""))
```

```{r}
# Calculating Southwest's Growth Factor from 2014 to 2015
southwest_growth = ((jun_southwest_passengers2015$TOTAL[2] / jun_southwest_passengers2015$TOTAL[1]) - 1) * 100
print(paste0("Southwest's Growth Rate (2014-2015): ", round(southwest_growth, digits = 2), "%", sep=""))
```

```{r}
# Calculating Southwest's Growth Factor from 2023 to 2024
southwest_growth2 = ((jun_southwest_passengers2024$TOTAL[2] / jun_southwest_passengers2024$TOTAL[1]) - 1) * 100
print(paste0("Southwest's Growth Rate (2023-2024): ", round(southwest_growth2, digits = 2), "%", sep=""))
```

### AIRLINE EVENTS

-   Frontier \@ 2006 : Holiday travel system problems
-   Frontier \@ 2014 : Customer anger over delays and cancellations, low quality score
-   JetBlue \@ 2005 : Redoing their business model after passengers voiced frustrations over paying high prices for fares and having flights delayed by slowdowns
-   Southwest \@ 2011 : Slowly recovering from 2008 recession
-   Spirit \@ 2013 : Lead in passenger complaints among the airline industry

### Analysis
```{r}
# Southwest has been in a steady linear decline since 2016 (ex-COVID) and is approaching an annual net income of break even. 
# Southwest with their acquisitions of AirTran in 2011 have increased their passengers and flights significantly. Alaska, Frontier, JetBlue, and Spirit have seen only modest growth in passengers and flights in the last 20 years.


# Despite the falling net income, operating revenue for the airlines has been increasing due to the increasing passenger numbers. All airlines, except JetBlue and Southwest, first started introducing checked bag fees in 2008, likely due to the Great Recession, in an attempt to recover revenue lost due to falling passenger numbers at that time. The overall trend in the airline industry over the last year, has been declining operating revenue which is causing airlines to look for additional sources of income.

# For Southwest, starting in the early 2000s, the number of passengers per flight was 75, and has steadily increased to the current average of approximately 120-125 passengers per flight. However, other airlines are able to get 150-160 passengers per flight hence generating more revenue per flight. Frontier is able to get 175 passengers per flight, yet still charge for checked bags and carry-ons, as well as seats. The other airlines also charge for checked bags and seats, which is becoming a revenue generating norm among the US airlines.

# The revenue per passenger for Alaska and JetBlue is significantly higher than Southwest and the other budget airlines. Alaska and JetBlue have additional revenue from their higher class fares and baggage fees, whereas Southwest has not captured that revenue yet and will be introducing those charges at the end of May to increase revenue.
```

### Comparison to JetBlue
```{r}
# JetBlue introduced baggage fees in 2015, and did not see any drop in operating revenue or in the number of passengers. By 2015, baggage fees have already been the norm in the US airline industry for 7 years, thus passengers have already been accustomed to paying those fees on every flight. Even though JetBlue had financial hardships from 2004-2010 due to rising fuel prices and other factors, their customer base remained loyal and rated the airline high in customer satisfaction.
```

### Prediction for Southwest
```{r}
# Southwest has a lower passenger per flight number compared to the other airlines, and a decreasing net income. The airline needs to do something to boost profitability. Southwest’s ticket prices are not the cheapest in the airline industry, and by charging for bags, the overall price to fly with Southwest is going to be even more expensive.

# Based on the facts above, and similar data from competing airlines, we predict that Southwest will not see any decline in passengers in June 2025 directly attributed to baggage fees the airline will introduce at the end of May. Since charging fees is the norm for all other airlines, Southwest will follow market trends (i.e., general decline in passenger volume felt throughout the industry). We predict that the added revenue from the fees will increase the net income of Southwest by June 2026.

```

## Estimate Found Using Reference Class Forecasting

By Assmaa Aziz

```{r}
# There are uncontrollable factors beyond these policy changes that could influence the number of passengers flying Southwest. Examples of economical factors are inflation, rising costs of needs, increased unemployment, and Trump's tariffs. Examples of political factors are Trump's strict immigration policies and his increase in deportations. 

#To account for these outside influences, we're using reference classes by comparing Southwest's passenger data to other similar airlines that began charging baggage fees.

#We're looking at other domestic U.S. airlines who are competitors to Southwest and therefore serve a similar customer base.

# JetBlue began charging for checked baggage in 2015.
# Spirit began charging for carry-on bags in 2010.
# Frontier began charging for carry-on bags in 2013.
# Alaska began charging for checked baggage in 2010.

#For each reference class, we are looking at the change in the number of passengers after the implementation of the new policy. If we notice the same trend among all airlines during the same time period, then we can conclude the change is likely due to outside factors that affect all airlines and not the policy change. However, if we see a different trend only in the airline with the policy change, we can conclude the effect is likely due to the policy change.

#We'll approximate the change in the number of passengers for each reference class.

#Based on how these changes affected these comparable airlines with similar policy changes, we'll adjust our Fermi estimate.
```

```{r}
passengers_data

passengers_data_refcl = passengers_data %>% filter(Year <= 2024) %>% filter(Airline %in% c('JetBlue', 'Frontier', 'Spirit', 'Alaska', 'Southwest')) %>% 
  dplyr::select(Year, Month, TOTAL, Airline, MonthName)
#passengers_data_refcl <- passengers_data_refcl[c("Year", "Month", "TOTAL", "Airline", "MonthName")]

passengers_data_refcl_summary = passengers_data_refcl %>% group_by(Airline, Year) %>% summarize(Total = sum(TOTAL))

plot_refcl = ggplot(passengers_data_refcl_summary, aes(x = Year, y = Total, color = Airline)) + geom_line() + geom_vline(xintercept = 2010, linetype = 'dashed') +  geom_vline(xintercept = 2013, linetype = 'dotted') +  geom_vline(xintercept = 2015, linetype = 'dotdash') + labs(title = 'Total Passengers Per Year')

plot_refcl
```

```{r}
#OBSERVATIONS

#Spirit, Alaska
#The policy change occurred in 2010. After 2010, both Spirit and Alaska airlines follow an upward trend. The remaining airlines also follow an upward trend for several years after 2010. The total number of passengers on domestic flights continued to increase for Spirit and Alaska after implementing these changes. This suggests the policy change did not impact passenger volume.

#Frontier
#The policy change occurred in 2013. After 2013, Frontier follows an upward trend. The remaining airlines follow an upward trend for several years after 2013. The total number of passengers on domestic flights continued to increase for Frontier after implementing these changes. This suggests the policy change did not impact passenger volume.

#JetBlue
#The policy change occurred in 2015. After 2015, JetBlue is continues increasing and then begins to flatten out and decline a couple years later, approximately around 2018. At around the same point, Alaska goes from increasing to flattening out, similar to JetBlue's trend in recent years. This suggests the influence of outside factors affecting both JetBlue and Alaska. However this trend is only visible in 1 airline unrelated to policy changes, which suggests the cause is likely something specific to Alaska airlines. The remaining airlines follow an upward trend for several years after 2015. JetBlue declining while other airlines increase suggests that JetBlue's change in baggage policy had a some impact. Since the decline is small and not a significant drop, it suggests the baggage policy had a small impact but ultimately did not dramatically decrease passenger volume.

#In all 3 time periods, Southwest has a significantly higher number of domestic passengers than our reference classes. After each policy change by other airlines, Southwest also experienced a dramatic increase. This suggests that some passengers may have switched over to Southwest after other airlines introduced new baggage fees.

#Overall, we didn't find strong evidence to suggest that implementing new baggage fees leads to a decline in passenger volume. In the budget airlines, Spirit and Frontier, their passenger volume continued to grow after implementing baggage fees. JetBlue experienced a very small decline after the policy change, but nothing large enough to suggest a significant drop in passenger volume. This suggests that implementing baggage fees has a small impact and doesn't hurt demand in the long-term.
```

```{r}
#Percent Change

#Spirit, Alaska 2010
before_2010 = passengers_data_refcl_summary %>% filter(Year == 2009 & (Airline == 'Spirit' | Airline == 'Alaska'))
after_2010 =  passengers_data_refcl_summary %>% filter(Year == 2011 & (Airline == 'Spirit' | Airline == 'Alaska'))

percent_change_2010 = ((after_2010$Total - before_2010$Total)/before_2010$Total)*100

percent_change_2010 = data.frame(airline = before_2010$Airline, percent_change_2010 = percent_change_2010)

percent_change_2010

#Frontier 2013
before_2013 = passengers_data_refcl_summary %>% filter(Year == 2012 & Airline == 'Frontier')
after_2013 = passengers_data_refcl_summary %>% filter(Year == 2014 & Airline == 'Frontier')

percent_change_2013 = ((after_2013$Total - before_2013$Total)/before_2013$Total)*100


percent_change_2013 = data.frame(airline = before_2013$Airline, percent_change_2013 = percent_change_2013)
percent_change_2013

#JetBlue 2015
before_2015 =  passengers_data_refcl_summary %>% filter(Year == 2014 & Airline == 'JetBlue')
after_2015 = passengers_data_refcl_summary %>% filter(Year == 2016 & Airline == 'JetBlue')

percent_change_2015 = ((after_2015$Total - before_2015$Total)/before_2015$Total)*100

percent_change_2015 = data.frame(airline = before_2015$Airline, percent_change_2015 = percent_change_2015)
percent_change_2015

```

```{r}
#2010
#After implementing their new policies, Alaska and Spirit's passenger volumes grew by 14.62% and 36.26% respectively.

#2013
#After implementing the new policies, Frontier's passenger volume grew by 13.09%.

#2015
#After implementing the new policies, JetBlue's passenger volume grew by 19.29%.
#This growth suggests baggage fees do not hurt consumer demand.

#BASE ESTIMATE
#Recall our forecasting question: Will changes to Southwest's policies in May 2025 decrease the total passengers in June 2025 and at the end of the third quarter?

#Because we found that all of our reference classes experienced growth after these policy changes, this suggests that the probability the total passengers will decrease from May 2025 to June 2025 is low. 

#Recall Laplace's rule of succession. If an event has has n opportunities to occur but has never happened, we assign probability 1/(n+2) to it happening the next time.

#We found that out of 4 instances of policy changes among different airlines, we didn't see a significant decrease in total passengers. Therefore, based on the results of our reference classes, our estimate of Southwest experiencing a decrease in total passengers after these policy changes is 1/(4+2) = 1/6 = 16.67%
ref_cl_est = 0.1667
```

##Reference Class Estimate Updated to Include Impact of Public Backlash

```{r}
#Another unpredictable factor that could impact our forecast is public backlash. Unlike our reference classes, Southwest's brand identity is tied to providing free checked bags, so it's possible public backlash could lead to a greater chance of a decrease in total passengers. For the reference classes we analyzed, their policy changes occurred before the rise of modern social media, so this factor likely didn't apply to them.

#Public backlash through social media could amplify negative sentiment and consumers could show their dissatisfaction for these policy changes by boycotting the airline.

#To account for the possible influence of public opinion, we'll simulate various levels of public backlash (low, medium, high) and assign probabilities to the likelihood we think each has of occurring. For each level, we'll assign the impact it could have on our base estimate, ref_cl_est. Taking the average of that, we'll have an updated estimate.
```

```{r}
#Our base estimate that there will be a decrease in total passengers from May 2025 to June 2025 for Southwest after the policy changes.
base_est = ref_cl_est

#Whether there will be a low, medium, or high level of public backlash.
levels = c('low', 'medium','high')

#Our belief of the probability each level of backlash occurs
#60% chance there is little to no backlash, 30% chance of moderate backlash, 10% chance of major backlash.
probs = c(0.60, 0.30, 0.10)
s = sample(levels, size = 10000, replace = TRUE, prob = probs)

#For each level of backlash, how much we expect it to increase the likelihood that total passengers will decrease after the policy changes.
expected_drop = c(low = 0, medium = 0.05, high = 0.1)
impact = expected_drop[s]
simulated_ests = base_est + impact

df = data.frame(simulated_estimates = simulated_ests)
ggplot(df, aes(x = simulated_estimates)) + geom_histogram()

#Our new estimate after running our simulation.
new_est = mean(simulated_ests)
new_est
```

## Identifying distribution using MLE

## By Ari Hadjiyianni

```{r}
# Use AIC and BIC to find the best fit for the data
library(fitdistrplus)
library(MASS)
```

```{r}

sw_passengers <- passengers_data %>% filter(Airline == "Southwest")
sw_passengers_june <- sw_passengers %>% 
  filter(Year >= 2003 & Year < 2020 | Year >= 2022 & Year != 2025, Month == 6) 
sw_passengers_june$TOTAL <- gsub("[\\,]", "", sw_passengers_june$TOTAL) 
sw_passengers_june$TOTAL<- as.numeric(sw_passengers_june$TOTAL)
sw_passengers_june <- sw_passengers_june[c("Year", "Month", "DOMESTIC", "INTERNATIONAL", "TOTAL")]

```

```{r}
mean = mean(sw_passengers_june$TOTAL)
sd = sd(sw_passengers_june$TOTAL) 

true_data <- ggplot(sw_passengers_june, aes(x = factor(Year), y = TOTAL/1000000)) + 
  geom_col() + 
  theme(axis.text.x = element_text(angle = 90)) + 
  labs(x = "Year", y = "Total (in millions)",
        title = "Total Southwest Passengers (ex-COVID)")

print(true_data)
```


```{r}
# fitting different distributions to find the best one
# normal, lognormal, poisson

# adding weights since more recent years will be more relevant, i.e, there were less travels as a whole in the early 2000

# using quadratic weights, linear is not "aggressive" enough and exponetial puts too much weight on recent years. 
suppressWarnings({
  

weights <- (1:length(total))^2
sum <-  sum(weights)

norm_weights <- weights/sum

total = sw_passengers_june$TOTAL

norm <- fitdist(total, "norm") # bell shaped data
lognorm <- fitdist(total, "lnorm") # skewed data 
poisson <- fitdist(total, "pois") #counts etc. 

aic <- rep(0, 3)
bic <- rep(0, 3)

list <- list(norm, lognorm, poisson)
for (i in 1:3){
  stat <- gofstat(list[[i]])
  aic[i] = stat$aic
  bic[i] = stat$bic
}

fit_df <- data.frame(
  distribution = c("norm", "lognorm", "possion"), 
  AIC = aic, 
  BIC = bic
)

min_aic = fit_df$distribution[fit_df$AIC == min(fit_df$AIC)]
min_bic = fit_df$distribution[fit_df$BIC == min(fit_df$BIC)]

# log norm has the smallest AIC and BIC so it is the best fit distribution. 

# use weights

log_data <- log(total)

mu = sum(norm_weights * log_data)

sigma <-sqrt(norm_weights *(log_data - mu)^2)
})
```

```{r}
# log_data
suppressWarnings({
  

df <- data.frame(
  lg_passengers = log_data, index=c(1:19))

model <- lm(lg_passengers ~ index, data = df, weights = norm_weights)

june_2025 <- data.frame(index = 21)

prediction <- predict(model, newdata = june_2025, se.fit = TRUE)

mu_log <- prediction$fit
se_log <- prediction$se.fit

mu_pred <-  exp(mu_log + sigma^2/2) #using expectation of lognorm
mean_mu_pred <-  mean(mu_pred)
variance <- se_log^2 + sigma^2
se <- sqrt(variance)

# simulations for confidence interval predictions in june 2025

set.seed(613)

n <- 10000

N_log <- rnorm(n, mu_log, se)

N_pred <-  exp(N_log + sigma^2/2) 

lower <-  quantile(N_pred, 0.025)
upper <- quantile(N_pred, 0.975)
})

print(paste0("95% Confidence Interval for June 2025 Passengers:", round(lower, digits = 0), " to ", round(upper, digits = 0)))
print(paste0("Mean of 95% confidence interval:", round(mean_mu_pred, digits = 0)))



```

```{r}
# Third quarter simulation 


sw_passengers_quarter <- sw_passengers %>% 
  filter(Year >= 2003 & Year < 2020 | Year >= 2022 & Year != 2025, Month %in% c(7, 8, 9)) 
sw_passengers_quarter$TOTAL <- gsub("[\\,]", "", sw_passengers_quarter$TOTAL) 
sw_passengers_quarter$TOTAL<- as.numeric(sw_passengers_quarter$TOTAL) 

sw_passengers_quarter <- sw_passengers_quarter %>% 
  group_by(Year) %>% 
  summarize(total = sum(TOTAL))

mean = mean(sw_passengers_quarter$total)
sd = sd(sw_passengers_quarter$total) 

true_data <- ggplot(sw_passengers_quarter, aes(x = factor(Year), y = total/1000000)) + 
  geom_col() + 
  theme(axis.text.x = element_text(angle = 90)) + 
  labs(x = "Year", 
        y = "Total Passengers (in millions)",
       title = "Q3 Southwest Passengers")
print(true_data)
```


```{r}
# fitting different distributions to find the best one
# normal, lognormal, poisson

# adding weights since more recent years will be more relevant, i.e, there were less travels as a whole in the early 2000

# using quadratic weights, linear is not "aggressive" enough and exponetial puts too much weight on recent years. 
suppressWarnings({
  
weights <- (1:length(sw_passengers_quarter$total))^2
sum <-  sum(weights)

norm_weights <- weights/sum

totals = sw_passengers_quarter$total

norm <- fitdist(totals, "norm") # bell shaped data
lognorm <- fitdist(totals, "lnorm") # skewed data 
poisson <- fitdist(totals, "pois") #counts etc. 

aic <- rep(0, 3)
bic <- rep(0, 3)

list <- list(norm, lognorm, poisson)
for (i in 1:3){
  stat <- gofstat(list[[i]])
  aic[i] = stat$aic
  bic[i] = stat$bic
}

fit_df <- data.frame(
  distribution = c("norm", "lognorm", "possion"), 
  AIC = aic, 
  BIC = bic
)

min_aic = fit_df$distribution[fit_df$AIC == min(fit_df$AIC)]
min_bic = fit_df$distribution[fit_df$BIC == min(fit_df$BIC)]

# log norm has the smallest AIC and BIC so it is the best fit distribution. 

# use weights

log_data <- log(totals)

mu = sum(norm_weights * log_data)

sigma <-sqrt(norm_weights *(log_data - mu)^2)

df <- data.frame(
  lg_passengers = log_data, 
  index = c(1:19)
#  index = c(1:17, 18:20)
)

model <- lm(lg_passengers ~ index, data = df, weights = norm_weights)

quarter_2025 <- data.frame(index = 21)

prediction <- predict(model, newdata = quarter_2025, se.fit = TRUE)

mu_log <- prediction$fit
se_log <- prediction$se.fit

mu_pred <-  exp(mu_log + sigma^2/2) #using expectation of lognorm
mean_mu_pred <-  mean(mu_pred)
variance <- se_log^2 + sigma^2
se <- sqrt(variance)

# simulations for confidence interval predictions in june 2025

set.seed(613)

n <- 10000

N_log <- rnorm(n, mu_log, se)

N_pred <-  exp(N_log + sigma^2/2) 

lower <-  quantile(N_pred, 0.025)
upper <- quantile(N_pred, 0.975)
})

print(paste0("95% Confidence Interval for third quarter 2025 Passengers: ", round(lower, digits = 0), " to ", round(upper, digits = 0), sep = ""))
print(paste0("Mean of 95% confidence interval: ", round(mean_mu_pred, digits = 0)))

```
